Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-08-18 grove-paper.tex
Use \autoref and macros \edge and \vertex

@@

@@

@@

This section introduces collaborative structure editing in Grove by example. 
We begin in Section~\ref{sub:Program Representation} by introducing basic edit actions
and the underlying graph representation of expressions with holes.
We will then discuss scenarios that arise in the collaborative setting in 
Section~\ref{sub:Collaborative Structure Editing}.

TODO: gray-scale unchanged parts of diagrams

@@

The editor state for each replica in Grove consists of an expression with holes.
For example, Figure~\ref{fig:A simple tree with a hole} shows the simple arithmetic expression \texttt{x * \hole}, which
has a hole on the right side of the plus operator.

Internally, Grove represents an expression like this as a directed graph with a distinguished root vertex.
Each vertex represents a term and is labeled with a globally unique identifier and a term constructor.
In this figure, vertices~2 and~4 have the constructors \texttt{times} and \texttt{var(x)}, respectively.
Each term constructor has an associated set of child positions.
The \texttt{*} constructor has positions for \texttt{left} and \texttt{right} children
and the \texttt{var} constructors have no child positions but takes an argument naming the particular variable referenced
(e.g., \texttt{var(x)} for the variable \texttt{x}).
The root vertex has a distinguished identifier and a constructor with a single child position.
For simplicity, we visually represent the root with a dot.

@@

% In this paper we represent variables and numeric constants
% as leaves and represent their names or values as
% arguments to those constructors (e.g., \texttt{2} in \texttt{num(2)}).
% (In TODO:REF, we discuss how to allow edits within these.)

@@

(though as we show in Section~\ref{sec:formalism} creating a vertex is implicit in edge creation)

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:Addition-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   "$\texttt{var(y)}_6$" [> "right", > "5"']
  }
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Addition-b} after adding vertex~6 and edge~5}
\label{fig:Addition-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Addition}
\label{fig:Addition}
\end{figure}

We model edits to the code as sequences of additions and deletions of edges to this graph.
For example, suppose the user adds \texttt{y} as the right child of vertex~2 in
Figure~\ref{fig:A simple tree with a hole}
so that the expression becomes \texttt{x * y}.
This corresponds to creating a vertex for the variable reference \texttt{y} and adding an edge
to that vertex from vertex~2.
Figure~\ref{fig:Addition} shows the result, where the new vertex has identifier~6.

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   {}
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_6$}]
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Addition-b} after deleting edge~5}
\label{fig:Change-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   "$\texttt{var(z)}_8$" [> "right", > "7"']
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 "$\texttt{var(y)}_6$"
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Change-a} after adding vertex~8 and edge~7}
\label{fig:Change-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Change}
\label{fig:Change}
\end{figure}

@@

Next suppose we want to replace \texttt{y} with \texttt{z} so that the code becomes \texttt{x * z}.
This is accomplished by first deleting edge~5.
This results in Figure~\ref{fig:Change-a}, which represents the expression \texttt{x * \hole{}}.
After that deletion, we add a new vertex to represent~\texttt{z} and an
edge from vertex~2 to that vertex.
This results in the bottom graph in Figure~\ref{fig:Change-b},
which represents the expression \texttt{x * z}.

@@

Note that vertex~3 continues to exist, and if it had any children, those children could remain connected to it.
In our system, once a vertex is created it is never deleted.
This allows further manipulations of those vertexes as shown later in this section.

@@

\begin{figure}
% TODO: naive add that creates a conflict
% \begin{tikzpicture}
% \path (0,0)       node {\includegraphics[scale=0.4]{graphviz/wrapping/1.dot}};
% \path (0cm,-4cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/2.dot}};
% \path (0cm,-8cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/3.dot}};
% \path (0cm,-12cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/4.dot}};
% \path (0cm,-16cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/5.dot}};
% \end{tikzpicture}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style]
};
\path (-1cm,-3cm) graph[graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Change-b} after deleting edge~1}
\label{fig:Wrapping-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"']
 }
};
\path (-1cm,-3cm) graph[graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Wrapping-a} after adding vertex~8 and edge~7}
\label{fig:Wrapping-b}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"'] -> {
   "$\texttt{*}_2$" [> "left", > "9"'] -> {
    "$\texttt{var(x)}_4$" [> "left", > "3"'],
    "$\texttt{var(z)}_6$" [> "right", > "5"']
   },
   {}
  }
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Wrapping-b} after adding edge~9}
\label{fig:Wrapping-c}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Wrapping}
\label{fig:Wrapping}
\end{figure}

TODO: dotted lines for deleted edges

@@

Before \marginpar{DEL?} further exploring how to merge changes from multiple users,
we want to point out a common idiom that we call \emph{wrapping}.
Consider the first graph in Figure~\ref{fig:Wrapping}, which
represents the code \texttt{$\lambda$ x . x}, and suppose
the user's cursor is in the body field of the lambda (i.e., around the variable
reference to \texttt{x} in the body of the lambda).

If the user requests that a plus be inserted,
the editor may notice that there is something already in that hole.
A naive editor could certainly add the plus by creating a new
edge from the body index of the lambda to a freshly created \texttt{plus} vertex.
This would result in a conflict between the \texttt{plus} and the existing \texttt{var}
as shown in Figure~REF:TODO.
A better user experience
would be interpret the request for a plus to mean to
take whatever (if anything) is in the cursor
and place it as a child of the plus.
This \emph{wraps} the existing \texttt{var} in the \texttt{plus}
and results in the code \texttt{$\lambda$ x . x + \hole}.
(This is similar to the behavior in other structure editors such as in CITE:TODO.)
Translating this into edge deletes and adds results in the sequence of graphs in Figure~\ref{fig:Wrapping}.
Note that this is a layer above

@@

\begin{figure}
% \centering
% \begin{tikzpicture}
% \path (0,0)    node {\includegraphics[scale=0.75]{graphviz/move/1.dot}};
% \path (0,-5cm) node {\includegraphics[scale=0.75]{graphviz/move/2.dot}};
% \path (0,-9cm) node {\includegraphics[scale=0.75]{graphviz/move/3.dot}};
% \end{tikzpicture}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"']
 }
};
\path (0cm,-3cm) graph [graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Wrapping-c} after deleting edge~9}
\label{fig:Move-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"'] -> {
   {},
   "$\texttt{*}_2$" [> "right", > "11"'] -> {
    "$\texttt{var(x)}_4$" [> "left", > "3"'],
    "$\texttt{var(z)}_6$" [> "right", > "5"']
   }
  }
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Move-a} after adding edge~11}
\label{fig:Move-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Moving vertex~2 from the left child to the right child of vertex~8}
\label{fig:Move}
\end{figure}

@@

For example, consider Figure~\ref{fig:Move}, which represents
the movement of \texttt{1 * 2} from the left child of \texttt{plus}
to its right child.  The top graph is the code before the move.
First, edge~7 is deleted, which results in the middle graph
where the rooted code is now \texttt{\hole{} + \hole},
and vertex~6 is orphaned.

Next edge~16 is added, which results in the bottom graph
where the rooted code is now \texttt{\hole{} + 1 * 2},
and vertex~6 is no longer orphaned.

@@

For example, suppose there are users, A and B,
that are synchronized with each other and
have the code \texttt{3 + 4}.
That is to say, they both have
the top graphs in Figure~\ref{fig:Editing Different Parts of the Code}.

Next, suppose user A changes the \texttt{3} to a \texttt{13},
and user B changes the \texttt{4} to a \texttt{14}.
Before transmitting their edits to each other,
user A thus has left graph and user B has the right graph.
For user A, this involves deleting edge~7 and creating edge~14.
For user B, this involves deleting edge~10 and creating edge~18.

@@

Note that aside from transmitting the edge deletions and additions,
no coordination between A and B is needed to coordinate the merge.
For example, suppose that A and B did not transmit their edits to each other
but to a third user, C, that has the original graph at the top of Figure~\ref{fig:Editing Different Parts of the Code}.
User C can replay the edits from A and B in any order and will always get
the graph at the bottom of Figure~\ref{fig:Editing Different Parts of the Code}.

Stated more formally, our edit model obeys commutativity
as shown in Figure~\ref{fig:Commutativity}.
For any two sequences of edits $\vec{\alpha}$ and $\vec{\alpha}^\prime$,
applying them in either order produces the same results.

@@

Consider Figure~\ref{fig:Editing Nested Parts of the Code}.
Alice and Bob both start with the top graph, which represents the code \texttt{2 * x + \hole{}}.
Alice changes the \texttt{2} to a \texttt{3}, which results in the left graph.
Bob moves the \texttt{2 * x} from the left of the plus to the right of the plus
by deleting edge~7 and adding edge~16, which results in the right graph.
Since vertex and edge identifiers are preserved by these,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the bottom graph.

@@

A consequence of this design is that we allow users to edit
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of edge~7 before making her
edits and before receiving Bob's adding of edge~16.
Her edits would then be within the code deleted by Bob.

@@

Finally, we address genuine conflicts between edits.
Consider the top graph of Figure~\ref{fig:Direct Conflicts}.
Suppose Alice inserts the number literal \texttt{2} into the body of the lambda
while Bob inserts the function application \texttt{f x} also into the body of the lambda.
These results in the left and right graphs of \ref{fig:Direct Conflicts} respectively.
In particular pay attention to the edges going out of the \texttt{body} index of node~3.
In the left graph it is edge~10, which points to a number literal,
and in the right graph it is edge~13, which points to a function application.

We merge these by including \emph{both} edges as edges going out of the \texttt{body} index
of node~3.
This results in the bottom graph of Figure~\ref{fig:Direct Conflicts}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index (i.e. the \texttt{body}
index of node~3).
In the graph representation, we flag this conflict by highlighting its edges
in red.
In the textual representation, we flag this conflict with the notation \texttt{\{ 2 | f x \}}.
This is a conflict that much eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete edge~10 and then edit the \texttt{x}
to be \texttt{x + 2}.
Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code can be run.\footnote{
It may be possible to develop evaluation models that allow these sorts of conflicts,
but that is beyond the scope of this paper.}

Compare this to what happens when there are conflicts in a traditional,
line/diff-based version control system (e.g., \texttt{git}, \texttt{mercurial}, \texttt{svn}, etc.).
Those systems annotate files with difference markers (e.g., sequences of \texttt{<}, \texttt{>}, or \texttt{=})
when a merge conflict happens.
These show alternate versions versions of the code, and it is up to the user
to replace those alternates with the desired merged result.

@@

An example of this is shown in Figure~\ref{fig:Multi-parent conflict},
which starts with the top graph, which represents the expression \texttt{\hole + (\hole + x)}.
Suppose Alice moves the variable \texttt{x} to the left-most hole.
This results in the left graph, which represents the expression \texttt{x + (\hole + \hole)}.
At the same time, suppose Bob moves the variable \texttt{x} to the right-most hole.
This results in the right graph, which represents the expression \texttt{\hole + (x + \hole)}.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes edge TODO and adds edge TODO, while Bob deletes edge TODO and adds edge TODO.
As explained in Subsection~TODO:REF, deletes are idempotent so both Alice and Bob deleting
edge TODO is not a problem.
However, both edge TODO (added by Alice) and edge TODO (added by Bob) point to the same vertex.
This generates the bottom graph, in which vertex TODO as two parents pointing to it.
We call this situation a multi-parent conflict.

@@

In a single user context, a cycle can appear when a user pastes
a vertex as a child of itself.
An example of this is shown in Figure~\ref{fig:Single-User Cycles},
where the user has pasted vertex~TODO as a child of vertex~TODO.

@@

For example, consider the situation shown in Figure~\ref{fig:Multi-User Cycles}.
The top graph represents code that is synchronized between Alice and Bob
and represents the code \texttt{(1 * \hole) * (2 * \hole)}.
In the left graph, Alice moves vertex~TODO to the \texttt{right}
index of vertex~TODO.
The resulting graph represents the code \texttt{(1 * (2 * \hole)) * \hole}.
In the right graph, Bob moves vertex~TODO to the \texttt{right} index of vertex~TODO.
The resulting graph represents the code \texttt{\hole * (2 * (1 * \hole))}.
On their own, neither edit creates a cycle.
However, when these edits are merged,
not only is a cycle created, but
the vertexes in the cycle

@@

The language we are modeling is shown in Figure~\ref{fig:Syntax as a grammar}.
Standard for a simply-typed lambda calculus except that
all terms have holes (represented by $\_$)
and the argument to a lambda is its own term.
The latter of which allows us to express lambdas with a hole for the bound variable.

>>>>

%%%%%%%%

@@

%%%%%%%%
% \autoref from hyperref
\renewcommand{\AMSautorefname}          {Equation}
\renewcommand{\appendixautorefname}     {Appendix}
\renewcommand{\chapterautorefname}      {Chapter}
\renewcommand{\equationautorefname}     {Equation}
\renewcommand{\FancyVerbLineautorefname}{Line}
\renewcommand{\figureautorefname}       {Figure}
\renewcommand{\footnoteautorefname}     {Footnote}
\renewcommand{\Hfootnoteautorefname}    {Footnote}
\renewcommand{\itemautorefname}         {Item}
\renewcommand{\Itemautorefname}         {Item}
\renewcommand{\pageautorefname}         {Page}
\renewcommand{\paragraphautorefname}    {Section}
\renewcommand{\partautorefname}         {Part}
\renewcommand{\sectionautorefname}      {Section}
\renewcommand{\subparagraphautorefname} {Section}
\renewcommand{\subsectionautorefname}   {Section}
\renewcommand{\subsubsectionautorefname}{Section}
\renewcommand{\tableautorefname}        {Table}
\renewcommand{\theoremautorefname}      {Theorem}

@@

\newcommand{\edge}[1]{Edge~#1}
\newcommand{\vertex}[1]{Vertex~#1}

@@

This section introduces collaborative structure editing in Grove by example. 
We begin in \autoref{sub:Program Representation} by introducing basic edit actions
and the underlying graph representation of expressions with holes.
We will then discuss scenarios that arise in the collaborative setting in 
\autoref{sub:Collaborative Structure Editing}.

TODO: gray-scale unchanged parts of diagrams; and dashes for removed parts?

@@

The editor state for each replica in Grove consists of an expression with holes.
For example, \autoref{fig:A simple tree with a hole} shows the simple arithmetic expression~\texttt{x * \hole}, which
has a hole on the right side of the plus operator.

Internally, Grove represents an expression like this as a directed graph with a distinguished root vertex.
Each vertex represents a term and is labeled with a globally unique identifier and a term constructor.
In this figure, \vertex{2} and \vertex{4} have the constructors \texttt{times} and \texttt{var(x)}, respectively.
Each term constructor has an associated set of child positions.
The~\texttt{*} constructor has positions for~\texttt{left} and~\texttt{right} children
and the~\texttt{var} constructors have no child positions but takes an argument naming the particular variable referenced
(e.g.,~\texttt{var(x)} for the variable~\texttt{x}).
The root vertex has a distinguished identifier and a constructor with a single child position.
For simplicity, we visually represent the root with a dot.

@@

% In this paper we represent variables and numeric constants
% as leaves and represent their names or values as
% arguments to those constructors (e.g.,~\texttt{2} in~\texttt{num(2)}).
% (In TODO:REF, we discuss how to allow edits within these.)

@@

(though as we show in \autoref{sec:formalism} creating a vertex is implicit in edge creation)

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:Addition-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   "$\texttt{var(y)}_6$" [> "right", > "5"']
  }
 }
};
\end{tikzpicture}
\caption{\autoref{fig:Addition-b} after adding \vertex{6} and \edge{5}}
\label{fig:Addition-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Addition}
\label{fig:Addition}
\end{figure}

We model edits to the code as sequences of additions and deletions of edges to this graph.
For example, suppose the user adds~\texttt{y} as the right child of \vertex{2} in
\autoref{fig:A simple tree with a hole}
so that the expression becomes \texttt{x * y}.
This corresponds to creating a vertex for the variable reference~\texttt{y} and adding an edge
to that vertex from \vertex{2}.
\autoref{fig:Addition} shows the result, where the new vertex has identifier~6.

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   {}
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_6$}]
};
\end{tikzpicture}
\caption{\autoref{fig:Addition-b} after deleting \edge{5}}
\label{fig:Change-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   "$\texttt{var(z)}_8$" [> "right", > "7"']
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 "$\texttt{var(y)}_6$"
};
\end{tikzpicture}
\caption{\autoref{fig:Change-a} after adding \vertex{8} and \edge{7}}
\label{fig:Change-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Change}
\label{fig:Change}
\end{figure}

@@

Next suppose we want to replace~\texttt{y} with~\texttt{z} so that the code becomes~\texttt{x * z}.
This is accomplished by first deleting \edge{5}.
This results in \autoref{fig:Change-a}, which represents the expression~\texttt{x * \hole{}}.
After that deletion, we add a new vertex to represent~\texttt{z} and an
edge from \vertex{2} to that vertex.
This results in the bottom graph in \autoref{fig:Change-b},
which represents the expression~\texttt{x * z}.

@@

Note that \vertex{3} continues to exist, and if it had any children, those children could remain connected to it.
In our system, once a vertex is created it is never deleted.
This allows further manipulations of those vertexes as shown later in this section.

@@

\begin{figure}
% TODO: naive add that creates a conflict
% \begin{tikzpicture}
% \path (0,0)       node {\includegraphics[scale=0.4]{graphviz/wrapping/1.dot}};
% \path (0cm,-4cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/2.dot}};
% \path (0cm,-8cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/3.dot}};
% \path (0cm,-12cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/4.dot}};
% \path (0cm,-16cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/5.dot}};
% \end{tikzpicture}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style]
};
\path (-1cm,-3cm) graph[graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{\autoref{fig:Change-b} after deleting \edge{1}}
\label{fig:Wrapping-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"']
 }
};
\path (-1cm,-3cm) graph[graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{\autoref{fig:Wrapping-a} after adding \vertex{8} and \edge{7}}
\label{fig:Wrapping-b}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"'] -> {
   "$\texttt{*}_2$" [> "left", > "9"'] -> {
    "$\texttt{var(x)}_4$" [> "left", > "3"'],
    "$\texttt{var(z)}_6$" [> "right", > "5"']
   },
   {}
  }
 }
};
\end{tikzpicture}
\caption{\autoref{fig:Wrapping-b} after adding \edge{9}}
\label{fig:Wrapping-c}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Wrapping}
\label{fig:Wrapping}
\end{figure}

@@

Before \marginpar{DEL?} further exploring how to merge changes from multiple users,
we want to point out a common idiom that we call \emph{wrapping}.
Consider the first graph in \autoref{fig:Wrapping}, which
represents the code~\texttt{$\lambda$ x . x}, and suppose
the user's cursor is in the body field of the lambda (i.e., around the variable
reference to~\texttt{x} in the body of the lambda).

If the user requests that a plus be inserted,
the editor may notice that there is something already in that hole.
A naive editor could certainly add the plus by creating a new
edge from the body index of the lambda to a freshly created \texttt{plus} vertex.
This would result in a conflict between the \texttt{plus} and the existing \texttt{var}
as shown in Figure~REF:TODO.
A better user experience
would be interpret the request for a plus to mean to
take whatever (if anything) is in the cursor
and place it as a child of the plus.
This \emph{wraps} the existing \texttt{var} in the \texttt{plus}
and results in the code~\texttt{$\lambda$ x . x + \hole}.
(This is similar to the behavior in other structure editors such as in CITE:TODO.)
Translating this into edge deletes and adds results in the sequence of graphs in \autoref{fig:Wrapping}.
Note that this is a layer above

@@

\begin{figure}
% \centering
% \begin{tikzpicture}
% \path (0,0)    node {\includegraphics[scale=0.75]{graphviz/move/1.dot}};
% \path (0,-5cm) node {\includegraphics[scale=0.75]{graphviz/move/2.dot}};
% \path (0,-9cm) node {\includegraphics[scale=0.75]{graphviz/move/3.dot}};
% \end{tikzpicture}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"']
 }
};
\path (0cm,-3cm) graph [graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{\autoref{fig:Wrapping-c} after deleting \edge{9}}
\label{fig:Move-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"'] -> {
   {},
   "$\texttt{*}_2$" [> "right", > "11"'] -> {
    "$\texttt{var(x)}_4$" [> "left", > "3"'],
    "$\texttt{var(z)}_6$" [> "right", > "5"']
   }
  }
 }
};
\end{tikzpicture}
\caption{\autoref{fig:Move-a} after adding \edge{11}}
\label{fig:Move-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Moving \vertex{2} from the left child to the right child of \vertex{8}}
\label{fig:Move}
\end{figure}

@@

For example, consider \autoref{fig:Move}, which represents
the movement of~\texttt{1 * 2} from the left child of \texttt{plus}
to its right child.  The top graph is the code before the move.
First, \edge{7} is deleted, which results in the middle graph
where the rooted code is now~\texttt{\hole{} + \hole},
and \vertex{6} is orphaned.

Next \edge{16} is added, which results in the bottom graph
where the rooted code is now~\texttt{\hole{} + 1 * 2},
and \vertex{6} is no longer orphaned.

@@

For example, suppose there are users, A and B,
that are synchronized with each other and
have the code~\texttt{3 + 4}.
That is to say, they both have
the top graphs in \autoref{fig:Editing Different Parts of the Code}.

Next, suppose user A changes the~\texttt{3} to a~\texttt{13},
and user B changes the~\texttt{4} to a~\texttt{14}.
Before transmitting their edits to each other,
user A thus has left graph and user B has the right graph.
For user A, this involves deleting \edge{7} and creating \edge{14}.
For user B, this involves deleting \edge{10} and creating \edge{18}.

@@

Note that aside from transmitting the edge deletions and additions,
no coordination between A and B is needed to coordinate the merge.
For example, suppose that A and B did not transmit their edits to each other
but to a third user, C, that has the original graph at the top of \autoref{fig:Editing Different Parts of the Code}.
User C can replay the edits from A and B in any order and will always get
the graph at the bottom of \autoref{fig:Editing Different Parts of the Code}.

Stated more formally, our edit model obeys commutativity
as shown in \autoref{fig:Commutativity}.
For any two sequences of edits $\vec{\alpha}$ and $\vec{\alpha}^\prime$,
applying them in either order produces the same results.

@@

Consider \autoref{fig:Editing Nested Parts of the Code}.
Alice and Bob both start with the top graph, which represents the code~\texttt{2 * x + \hole{}}.
Alice changes the~\texttt{2} to a~\texttt{3}, which results in the left graph.
Bob moves the~\texttt{2 * x} from the left of the plus to the right of the plus
by deleting \edge{7} and adding \edge{16}, which results in the right graph.
Since vertex and edge identifiers are preserved by these,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the bottom graph.

@@

A consequence of this design is that we allow users to edit
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of \edge{7} before making her
edits and before receiving Bob's adding of \edge{16}.
Her edits would then be within the code deleted by Bob.

@@

Finally, we address genuine conflicts between edits.
Consider the top graph of \autoref{fig:Direct Conflicts}.
Suppose Alice inserts the number literal~\texttt{2} into the body of the lambda
while Bob inserts the function application~\texttt{f x} also into the body of the lambda.
These results in the left and right graphs of \autoref{fig:Direct Conflicts} respectively.
In particular pay attention to the edges going out of the \texttt{body} index of \vertex{3}.
In the left graph it is \edge{10}, which points to a number literal,
and in the right graph it is \edge{13}, which points to a function application.

We merge these by including \emph{both} edges as edges going out of the \texttt{body} index
of \vertex{3}.
This results in the bottom graph of \autoref{fig:Direct Conflicts}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index (i.e. the \texttt{body}
index of \vertex{3}).
In the graph representation, we flag this conflict by highlighting its edges
in red.
In the textual representation, we flag this conflict with the notation~\texttt{\{ 2 | f x \}}.
This is a conflict that much eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete \edge{10} and then edit the~\texttt{x}
to be~\texttt{x + 2}.
Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code can be run.\footnote{
It may be possible to develop evaluation models that allow these sorts of conflicts,
but that is beyond the scope of this paper.}

Compare this to what happens when there are conflicts in a traditional,
line/diff-based version control system (e.g., \texttt{git}, \texttt{mercurial}, \texttt{svn}, etc.).
Those systems annotate files with difference markers (e.g., sequences of~\texttt{<},~\texttt{>}, or~\texttt{=})
when a merge conflict happens.
These show alternate versions versions of the code, and it is up to the user
to replace those alternates with the desired merged result.

@@

An example of this is shown in \autoref{fig:Multi-parent conflict},
which starts with the top graph, which represents the expression~\texttt{\hole + (\hole + x)}.
Suppose Alice moves the variable~\texttt{x} to the left-most hole.
This results in the left graph, which represents the expression~\texttt{x + (\hole + \hole)}.
At the same time, suppose Bob moves the variable~\texttt{x} to the right-most hole.
This results in the right graph, which represents the expression~\texttt{\hole + (x + \hole)}.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes edge TODO and adds edge TODO, while Bob deletes edge TODO and adds edge TODO.
As explained in Subsection~TODO:REF, deletes are idempotent so both Alice and Bob deleting
edge TODO is not a problem.
However, both edge TODO (added by Alice) and edge TODO (added by Bob) point to the same vertex.
This generates the bottom graph, in which vertex TODO as two parents pointing to it.
We call this situation a multi-parent conflict.

@@

In a single user context, a cycle can appear when a user pastes
a vertex as a child of itself.
An example of this is shown in \autoref{fig:Single-User Cycles},
where the user has pasted vertex~TODO as a child of vertex~TODO.

@@

For example, consider the situation shown in \autoref{fig:Multi-User Cycles}.
The top graph represents code that is synchronized between Alice and Bob
and represents the code~\texttt{(1 * \hole) * (2 * \hole)}.
In the left graph, Alice moves vertex~TODO to the \texttt{right}
index of vertex~TODO.
The resulting graph represents the code~\texttt{(1 * (2 * \hole)) * \hole}.
In the right graph, Bob moves vertex~TODO to the \texttt{right} index of vertex~TODO.
The resulting graph represents the code~\texttt{\hole * (2 * (1 * \hole))}.
On their own, neither edit creates a cycle.
However, when these edits are merged,
not only is a cycle created, but
the vertexes in the cycle

@@

The language we are modeling is shown in \autoref{fig:Syntax as a grammar}.
Standard for a simply-typed lambda calculus except that
all terms have holes (represented by $\_$)
and the argument to a lambda is its own term.
The latter of which allows us to express lambdas with a hole for the bound variable.
